Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-05 inference.tex
Inference rules changed to include F_{ctxt}

At the core of our approach is a simple language ($\SL$) to express
set transformations (see Fig.~\ref{fig:logic-syntax}). The language
admits only set expressions that include variables ($x$), literals of
the form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $\stg$ satisfying proposition $\phi$ in a
set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
conditional set expression where the condition is a propositional
formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
are also variables in $\SL$, but are used to denote local and database
states, respectively. Constant sets can be written using set
literal expressions. For example, the set $\{1,2\}$ can be written as
$\{x \,|\, x=1 \vee x=2\}$. The language is carefully chosen to be
expressive enough to capture the semantics of $\txnimp$ statements (as
well as SQL operations more generally), yet simple enough to have a
semantics-preserving translation amenable for automated verification.

@@

transaction $\C{TXN}_i$.  Inference depends on the $\I$-constrained
rely relation $\R$, and the high-level invariant $I$ for reasons
that will become clear shortly.  Inference rules compute, for each
command $c$, a (meta) function $\F$ that returns a set as an
expression in $\SL$, given a pair of sets ($\stl$ and $\stg$) that
describe local and a global database states, respectively. The
expression returned by $\F(\stl,\stg)$ (abstractly) describes the set
of records that get added to $\stl$ as a result of executing $c$ under
$\stl$ and $\stg$.  Thus, $\F$ captures the \emph{effect} part of the
state transformer of $c$, which is the function
$\lambda(\stl,\stg).~\stl \cup \F(\stl,\stg)$\footnote{Recall that the
  operational semantics treats deletion of records as the addition of
  the deleted record with its \C{del} field set to true in the local
  store.}. For $\F$ to be useful in RG verification, it needs to be
stable w.r.t the rely relation $\R$. The stability condition on
effects can be defined thus:

@@

  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg',\nubar.~
  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')

@@

possible because of how the inference rules are structured.
Intuitively, the stability condition requires an effect to describe
the same set of records before and after an interference. The
equivalence in $\SL$ translates to equality in first-order logic, as
we describe below. In the inference rules, stability is enforced

@@

an effect and returns a new effect that is guaranteed to be stable
under $\R$.  $\stabilize{\cdot}$ achieves the stability guarantee by
abstracting away the bound global state ($\stg$) in an unstable $\F$
to an existentially bound $\stg'$ as described below:

@@

  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F).\\
  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) 

@@

the existentially bound $\stg'$. The example in

@@

a \C{select1} query, whose effect ($\F$) is essentially a singleton

@@

$\F(\stl,\stg) = \{r \,|\, r\in\stg \wedge r.\C{id} = \C{acc\_id}\}$).
However, this effect is unstable because $\F(\stl,\stg')$ may not be
the same set as $\F(\stl,\stg)$ when $\stg'\neq\stg$. A record

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of
$c_1$ and $c_2$, respectively, except that $\F_2$ is applied to a
state ($\stl$) to which $\F_1$ has already been applied, reflecting
their order of reduction. The inference rule for \C{FOREACH} takes
advantage of the $\SL$'s bind operator to lift the effect inferred for
the loop body to the level of the loop. Since records added to $\stl$
in each iteration of \C{FOREACH} are independent of the previous
iteration (recall that we make a local context independence assumption
about database programs; Sec. ~\ref{sec:opsem}), sequential
composition of the effects of different iterations is the same as their
parallel composition. Since the loop body is executed once per each
$z\in x$, the effect of the the loop is a union of effects ($\F$)
for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

@@

\subsection{Soundness of Inference}

We can now establish a correspondence between the inference rules given
above and the RG judgment of \S\ref{sec:reasoning}:
\begin{theorem}
  \label{thm:inference-sound}
  Let the set expression $\F(\emptyset,\stg)$ have a natural interpretation
  in the theory of sets.  Then, for all $i$,$R$,$I$,$c$,$\F$,
  if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}

\paragraph{{\sc Proof Sketch.}} We prove a slightly stronger statement :
For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
  \begin{smathpar}
    \begin{array}{c}
    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
    I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
    \end{array}
  \end{smathpar}
The proof follows by structural induction on $c$. Assume that $P
\Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$ and $Q
\Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For each type
of $c$, our strategy is to show the premises of the corresponding
RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
statements. In all these cases, the form of $\F$ directly corresponds
to the condition linking the pre-condition $P$ and post-condition $Q$
in the premises of the RG-rules. If $\stable(\R, \F)$, then
$\stabilize{\F} = \F$ and hence the proof directly follows. Otherwise,
$\stabilize{\F} = \lambda (\stl, \stg).\existsl(\stg', I(\stg'),
\F(\stl, \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
in the above set expression, and hence $\F$ applied on that state
matches the condition linking $P$ and $Q$. The stability of $P$ with
respect to $\R$ follows directly from the stability of $I$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
   \stable(R,\I)\spc
   \stable(R,I)\\
   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc  
   c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \spc \stable(\R_c,\F)\\
   \forall \stl,\stg.~ \stl = F(\emptyset,\stg) \Rightarrow 
   G(\stg, \stl\gg\stg)\spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
% \RULE
% {
%   \stable(R,\I)\spc
%   \stable(R,I)\spc
%   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
%   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
%    \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
%   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
%     G(\stg, \stl \rhd \stg)\spc
%   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
% }
% {
%   \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
% }
% \end{array}
% \end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $G$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

>>>>

At the core of our approach is a simple language ($\SL$) to express
set transformations (see Fig.~\ref{fig:logic-syntax}). The language
admits only set expressions that include variables ($x$), literals of
the form $\{x \,|\, \varphi\}$ where $\varphi$ is a propositional
(quantifier-free) formula on $x$, a restricted form of existential
quantification that binds a set $\stg$ satisfying proposition $\phi$ in a
set expression $s$, a monadic composition of two set expressions
($s_1$ and $s_2$) composed using a bind ($\bind$) operation, a
conditional set expression where the condition is a propositional
formula, and a union of two set expressions. Symbols $\stl$ and $\stg$
are also variables in $\SL$, but are used to denote local and database
states, respectively. Constant sets can be written using set
literal expressions. For example, the set $\{1,2\}$ can be written as
$\{x \,|\, x=1 \vee x=2\}$. The language is carefully chosen to be
expressive enough to capture the semantics of $\txnimp$ statements (as
well as SQL operations more generally), yet simple enough to have a
semantics-preserving translation amenable for automated verification.

@@

transaction $\C{TXN}_i$. The rules compute, for each command $c$, a
(meta) function $\F$ that returns a set of records as an expression in
$\SL$, given a global database $\stg$. Intuitively, $\F(\stg)$
abstracts the set of records added to the local database $\stl$ as a
result of executing $c$ under $\stg$ (i.e., $\stg \vdash
(\tbox{c}_i,\stl) \stepsto^{*}_{R} (\tbox{\cskip}_i, \stl \cup
\F(\stg))$)\footnote{Recall that the operational semantics treats 
  deletion of records as the addition of the deleted record with its
  \C{del} field set to true in the local store.}. Note that $\F$
actually captures the \emph{effect} part of the state transformer,
but we nonetheless refer to it as a transformer. Since the execution
is subject to isolation-constrained interference, the inference
judgment depends on the isolation-constrained rely relation $\R$,
which is used to enforce the stability of the state transformer $\F$.
Recall that $\R$ is a tri-state rely relation over $\stl$, $\stg$ and
$\stg'$, that admits an interference from $\stg$ and $\stg'$ depending
on the local database state $\stl$. Thus, the stability of the state
transformer $\F$ of $c$ with respect to $\R$ needs to take into
account the (possible) prior state of the local database $\stl$, which
depends on the context (sequence of previous commands) of $c$, and
computed by the corresponding state transformer $\Fx$. Thus, the
semantics of the state transformer can be understood in terms of the
RG judgment as following (formalized as
Theorem~\ref{thm:inference-sound} in Sec.~\ref{sec:inference-sound}):
  \begin{array}{c}
    \R \vdash \hoare{\lambda(\stl,\stg).~ \stl = \Fx(\stg)}{c}
    {\lambda(\stl,\stg).~ \stl = \Fx(\stg) \cup \F(\stg)}
In the above RG judgment, let us call the pre-condition $P$, and
post-condition $Q$. The stability condition on the state transformer
$\F$ can be derived from the stability condition on $Q$. Observe that
for $Q$ to be stable, $\Fx(\stg') \cup \F(\stg')$ must be equal to
$\Fx(\stg) \cup \F(\stg)$, where $\stg$ and $\stg')$ are related by
$R$ (ignore $\I$ for the moment). Assuming that $P$ is stable,
$\Fx(\stg')=\Fx(\stg)$ is already given, leaving $\F(\stg')=\F(\stg)$
to be enforced. Thus, the stability of $\F$ in in the context of $\Fx$
(written $\inctxt{\Fx}{\F}$) is defined as following:

@@

  \stable(\R,\inctxt{\Fx}{\F}) & \Leftrightarrow & \forall \stg,\stg',\nubar.~
  \R(\Fx(\stg) \cup \F(\stg),\stg,\stg') \Rightarrow \F(\stg) = \F(\stg')

@@

possible because of how the inference rules are structured. The
equality in $\SL$ translates to equivalence in first-order logic, as
we describe later. In the inference rules, stability is enforced

@@

a transformer $\F$ (in its context $\Fx$) and returns a new
transformer that is guaranteed to be stable under $\R$.
$\stabilize{\cdot}$ achieves the stability guarantee by abstracting
away the bound global state ($\stg$) in an unstable $\F$ to an
existentially bound $\stg'$ as described below:

@@

  \stabilize{\inctxt{\Fx}{\F}} & = & \F & \texttt{if }
  \stable(\R,\inctxt{\Fx}{\F}).\\
  & = & \lambda (\stg).~\existsl(\stg',I(\stg'),\F(\stg')) 

@@

an existentially bound $\stg'$. The example in

@@

a \C{select1} query, whose transformer $\F$ is essentially a singleton

@@

$\F(\stg) = \{r \,|\, r\in\stg \wedge r.\C{id} = \C{acc\_id}\}$).
However, $\F$ is unstable because $\F(\stg')$ may not be
the same set as $\F(\stg)$ when $\stg'\neq\stg$. A record

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of
$c_1$ and $c_2$, respectively, except that $\F_2$ is computed in a
context that includes $\F_1$. The inference rule for \C{FOREACH} takes
advantage of the $\SL$'s bind operator to lift the effect inferred for
the loop body to the level of the loop. Since records added to $\stl$
in each iteration of \C{FOREACH} are independent of the previous
iteration (recall that we make a local context independence assumption
about database programs; Sec. ~\ref{sec:opsem}), sequential
composition of the effects of different iterations is the same as
their parallel composition. Since the loop body is executed once per
each $z\in x$, the effect of the the loop is a union of effects ($\F$)
for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
That is, $\F_{loop}(\stg) = \bigcup_{z\in
x}\F_{body}(\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stg) = x \bind (\lambda
z.~F_{body}(\stg))$, which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

@@

\subsection{Soundness of Inference}
\label{sec:inference-sound}

We now formally state the correspondence between the inference rules
given above and the RG judgment of \S\ref{sec:reasoning}:
\begin{theorem}
  \label{thm:inference-sound}
  For all $i$,$R$,$I$,$c$,$\Fx$, $\F$, if $\stable(\R,I)$ and $\Fx
  \vdash c \elabsto \F$, then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\Fx(\stg) \conj
    I(\stg)}{c}{\lambda(\stl,\stg).~\stl = \Fx(\stg) \cup \F(\stg)
    \conj I(\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}

\paragraph{{\sc Proof Sketch.}} 
% We prove a slightly stronger statement :
% For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
%   \begin{smathpar}
%     \begin{array}{c}
%     \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
%     I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
%     \end{array}
%   \end{smathpar}
The proof follows by structural induction on $c$. Assume that $P
\Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$ and $Q
\Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For each type
of $c$, our strategy is to show the premises of the corresponding
RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
statements. In all these cases, the form of $\F$ directly corresponds
to the condition linking the pre-condition $P$ and post-condition $Q$
in the premises of the RG-rules. If $\stable(\R, \F)$, then
$\stabilize{\F} = \F$ and hence the proof directly follows. Otherwise,
$\stabilize{\F} = \lambda (\stg).\existsl(\stg', I(\stg'),
\F( \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
in the above set expression, and hence $\F$ applied on that state
matches the condition linking $P$ and $Q$. The stability of $P$ with
respect to $\R$ follows directly from the stability of $I$.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below ($\Fempty =
\lambda(\stg).~\emptyset$ denotes an empty context):
\begin{smathpar}
\begin{array}{c}
\RULE
{
   \stable(R,\I)\spc
   \stable(R,I)\\
   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc
   \Fempty \vdash c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \\
   \stable(\R_c,\inctxt{\Fempty}{\F}) \spc
   \forall \stg.~ G(\stg,\F(\stg)) \spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
% \RULE
% {
%   \stable(R,\I)\spc
%   \stable(R,I)\spc
%   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
%   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
%    \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
%   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
%     G(\stg, \stl \rhd \stg)\spc
%   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
% }
% {
%   \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
% }
% \end{array}
% \end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $G$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.
